Preprocessing

### 2024-08-12 nbs/guides/preprocessing/preprocessing.qmd
docs(preprocessing): initialize preprocessing guide
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Preprocessing
execute: 
  freeze: true
  eval: true
  warning: false
  error: false
  cache: true
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: katex
  ipynb: default
---

@@

>>>>

---
title: Preprocessing
execute: 
  freeze: false
  eval: true
  warning: false
  error: false
  cache: true
toc: true
number-sections: true
highlight-style: gruvbox
csl: ../../bibstyle.csl
lightbox: auto
format:
  html:
    html-math-method: katex
  ipynb: default
---

@@

## Download data

::: {#wrn-downsampled-data .callout-warning collapse=true title="Downsampled Data"}
In this notebook we illustrate how to download and review results using
unrealistic downsampled data. Therefore, this notebook should not be taken
to represent the results of applying the model to real data.
:::

We download an example data set to be used for illustration of the 
preprocessing steps.

::: {#wrn-mutable-data .callout-warning collapse=true title="Mutable objects"}
Ideally we would work with immutable data structures to avoid losing
information about how a given data object has been transformed. However,
the usual object-oriented approach relies on repeated mutation of the
same object. The only defense against confusion we have here is to review 
the updated properties of the underlying object after applying effectful
methods or functions that have the ability to mutate an object's state in
any way. This is why we use the `pyrovelocity.utils.anndata_string` function
after each step that may mutate the data object. Of course, without hashing
each subcomponent we will still not be able to verify that pre-existing
components of the object were not modified but at least we can verify
the relationship between added and removed components this way.
:::

```{python}
# | label: download-data
from pyrovelocity.tasks.data import download_dataset
from pyrovelocity.tasks.data import load_anndata_from_path
from pyrovelocity.utils import anndata_string, print_string_diff

dataset_path = download_dataset(data_set_name="pancreas")
adata = load_anndata_from_path(dataset_path)
initial_data_state_representation = anndata_string(adata)
print(initial_data_state_representation)
```

## Preprocessing steps

### Subset data

We will eventually scale up the analysis to the full data set, but it is
helpful to run experiments on downsampled data to speed up iteration
prior to running on the full data set. Here we downsample the data to
300 observations and 200 variables. Some steps may benefit from even
further downsampling but this is often a good starting point.

```{python}
# | label: subset-data
import os
import numpy as np
from pyrovelocity.utils import str_to_bool

PYROVELOCITY_DATA_SUBSET = str_to_bool(
    os.getenv("PYROVELOCITY_DATA_SUBSET", "True")
)

if PYROVELOCITY_DATA_SUBSET:
    adata = adata[
        np.random.choice(
            adata.obs.index,
            size=300,
            replace=False,
        ),
        :,
    ].copy()
subset_anndata_representation = anndata_string(adata)
print_string_diff(
    text1=initial_data_state_representation,
    text2=subset_anndata_representation,
    diff_title="Downsampled data diff",
)
```

You can save the downsampled data for later use, but for small data sets
less than $O(10^4)$ observations it is not too expensive to reload the
full data set if the identity of the random sample does not need to be reproducible.

We will make use of the raw count data in constructing a probabilistic model
of the spliced and unspliced read or molecule counts. The most commonly used
preprocessing methods overwrite this information with normalized or transformed
versions so we always begin by saving a copy of the count data.

### Save count data

```{python}
# | label: save-count-data
from pyrovelocity.tasks.preprocess import copy_raw_counts

copy_raw_counts(adata)
copied_raw_counts_representation = anndata_string(adata)
print_string_diff(
    text1=subset_anndata_representation,
    text2=copied_raw_counts_representation,
    diff_title="Copy raw counts diff",
)
```

### Pre-filtration QC

```{python}
# | label: pre-filtration-quality-control-metrics
import scanpy as sc

adata.var["mt"] = adata.var_names.str.startswith("MT-" or "Mt-" or "mt-")
adata.var["ribo"] = adata.var_names.str.startswith(
    "RPS" or "Rps" or "rps" or "RPL" or "Rpl" or "rpl"
)

sc.pp.calculate_qc_metrics(
    adata=adata,
    qc_vars=["mt", "ribo"],
    percent_top=None,
    log1p=False,
    inplace=True,
)
qc_metrics_representation = anndata_string(adata)
print_string_diff(
    text1=copied_raw_counts_representation,
    text2=qc_metrics_representation,
    diff_title="Quality control metrics diff",
)
```

```{python}
# | label: pre-filtration-plot-genes-by-counts
import seaborn as sns
import pandas as pd

numeric_obs = adata.obs.copy()
numeric_obs["n_genes_by_counts"] = pd.to_numeric(
    numeric_obs["n_genes_by_counts"],
    errors="coerce",
)
qc_counts = sns.jointplot(
    data=numeric_obs,
    x="total_counts",
    y="n_genes_by_counts",
    color="#ff6a14",
    marginal_ticks=True,
    kind="scatter",
    alpha=0.4,
)
qc_counts.plot_joint(
    sns.kdeplot,
    color="gray",
    alpha=0.6,
)
qc_counts.set_axis_labels(
    xlabel="Total counts in cell",
    ylabel="Number of genes >=1 count in cell",
)
```

### Compute cytotrace statistics

For some data sets we apply cytotrace as a proxy for progression through 
lineage-specific differentiation.

```{python}
# | label: cytotrace
from pyrovelocity.analysis.cytotrace import cytotrace_sparse

cytotrace_sparse(adata, layer="spliced")
cytotrace_representation = anndata_string(adata)
print_string_diff(
    text1=copied_raw_counts_representation,
    text2=cytotrace_representation,
    diff_title="Cytotrace diff",
)
```

### Filter and normalize data

Next we use the preprocessing functions built-in to the scvelo package to 
filter genes that have at least a given minimum number of counts,

```{python}
# | label: filter-genes-by-counts
import scvelo as scv

scv.pp.filter_genes(
    adata,
    min_shared_counts=30,
)
filter_genes_representation = anndata_string(adata)
print_string_diff(
    text1=cytotrace_representation,
    text2=filter_genes_representation,
    diff_title="Filter genes diff",
)
```

normalize the data,

```{python}
# | label: normalize-per-cell
scv.pp.normalize_per_cell(adata)
normalize_representation = anndata_string(adata)
print_string_diff(
    text1=filter_genes_representation,
    text2=normalize_representation,
    diff_title="Normalize per cell diff",
)
```

and filter the genes by their degree of dispersion.

```{python}
# | label: filter-genes-by-dispersion
scv.pp.filter_genes_dispersion(
    adata,
    n_top_genes=2000,
)
filter_dispersion_representation = anndata_string(adata)
print_string_diff(
    text1=normalize_representation,
    text2=filter_dispersion_representation,
    diff_title="Filter dispersion diff",
)
```

Aside from the observation-level subsampling these operations are destructive 
and so you would need to reload the data to recover its original state after
attempting to apply them.

Finally we use the scanpy package to log-transform the data.
Note again we've retained the count data and are only doing this as it is
expected by a variety of downstream analysis tools.

```{python}
# | label: log-transform-data-matrix
import scanpy as sc

sc.pp.log1p(adata)
preprocessed_data_state_representation = anndata_string(adata)
print_string_diff(
    text1=filter_dispersion_representation,
    text2=preprocessed_data_state_representation,
    diff_title="Log transform diff",
)
```

Having executed this series of functions we've added or updated the following
components relative to the initial data state

```{python}
# | label: print-data-state-summary
print_string_diff(
    text1=initial_data_state_representation,
    text2=preprocessed_data_state_representation,
    diff_title="Filtration and normalization summary diff",
    diff_context_lines=5,
)
```

### Compute nearest neighbors graph

We do not make direct use of nearest neighbor graphs, and some data sets
come with one precomputed, but we recompute the graph
for cross-compatability with other tools that may require it.

```{python}
# | label: compute-nearest-neighbors-graph
sc.pp.neighbors(
    adata=adata,
    n_pcs=30,
    n_neighbors=30,
)
nearest_neighbors_representation = anndata_string(adata)
print_string_diff(
    text1=preprocessed_data_state_representation,
    text2=nearest_neighbors_representation,
    diff_title="Nearest neighbors diff",
)
```

The nearest neighbor graph is used in particular for nearest-neighbor-based 
averaging of gene expression

```{python}
# | label: compute-nearest-neighbor-averages
scv.pp.moments(
    data=adata,
    n_pcs=30,
    n_neighbors=30,
)
nearest_neighbor_averaged_representation = anndata_string(adata)
print_string_diff(
    text1=nearest_neighbors_representation,
    text2=nearest_neighbor_averaged_representation,
    diff_title="Nearest neighbor averaged diff",
)
```

### Estimate expression dynamics

We compute the estimates of the expression dynamics parameters using the EM-based
maximum-likelihood algorithm from the scvelo package

```{python}
# | label: compute-em-ml-velocity
scv.tl.recover_dynamics(
    data=adata,
    n_jobs=-1,
    use_raw=False,
)
em_ml_dynamics_representation = anndata_string(adata)
print_string_diff(
    text1=nearest_neighbor_averaged_representation,
    text2=em_ml_dynamics_representation,
    diff_title="EM ML dynamics diff",
)
```

and use these to estimate the gene-specific velocities.

```{python}
# | label: compute-gene-specific-velocities
scv.tl.velocity(
    data=adata,
    mode="dynamical",
    use_raw=False,
)
em_ml_velocities_representation = anndata_string(adata)
print_string_diff(
    text1=em_ml_dynamics_representation,
    text2=em_ml_velocities_representation,
    diff_title="EM ML velocities diff",
)
```

### Compute leiden clusters

We add leiden clustering to the data set for algorithms that work with 
unlabeled clusters.

```{python}
# | label: leiden-clustering
sc.tl.leiden(adata=adata)
leiden_representation = anndata_string(adata)
print_string_diff(
    text1=em_ml_velocities_representation,
    text2=leiden_representation,
    diff_title="Leiden clustering diff",
)
```

### Subset genes

We also subset the genes to a smaller number to enable faster illustration of
model training and inference even on laptops with significantly limited resources.

```{python}
# | label: subset-genes
likelihood_sorted_genes = (
    adata.var["fit_likelihood"].sort_values(ascending=False).index
)

if PYROVELOCITY_DATA_SUBSET:
    adata = adata[:, likelihood_sorted_genes[:200]].copy()

subset_genes_representation = anndata_string(adata)
print_string_diff(
    text1=leiden_representation,
    text2=subset_genes_representation,
    diff_title="Subset genes diff",
)
```

### Compute velocity graph

We compute the velocity graph based on cosine similarity of the gene-specific velocities
estimated by the scvelo package.

```{python}
# | label: compute-velocity-graph
scv.tl.velocity_graph(
    data=adata,
    n_jobs=-1,
)
velocity_graph_representation = anndata_string(adata)
print_string_diff(
    text1=subset_genes_representation,
    text2=velocity_graph_representation,
    diff_title="Velocity graph diff",
)
```

### Compute velocity embedding

We compute the velocity embedding using the velocity graph and the gene-specific velocities.

```{python}
# | label: compute-velocity-embedding
scv.tl.velocity_embedding(
    data=adata,
    basis="umap",
)
velocity_embedding_representation = anndata_string(adata)
print_string_diff(
    text1=velocity_graph_representation,
    text2=velocity_embedding_representation,
    diff_title="Velocity embedding diff",
)
```

### Compute gene-wise latent time

We compute the gene-specific latent time estimated by the scvelo package.

```{python}
# | label: compute-latent-time
scv.tl.latent_time(
    data=adata,
)
latent_time_representation = anndata_string(adata)
print_string_diff(
    text1=velocity_embedding_representation,
    text2=latent_time_representation,
    diff_title="Latent time diff",
)
```

### Summarize preprocessing

We can finally summarize the changes induced by the preprocessing steps 
applied to the data set.

```{python}
# | label: summarize-preprocessing
print_string_diff(
    text1=initial_data_state_representation,
    text2=latent_time_representation,
    diff_title="Preprocessing summary diff",
    diff_context_lines=5,
)
```

## Plots

### Data quality

We plot the two-dimensional density of genes with greater than or equal to one count in a given cell versus the total counts in that cell for the data set.

```{python}
# | label: post-filtration-quality-control-metrics
sc.pp.calculate_qc_metrics(
    adata=adata,
    qc_vars=["mt", "ribo"],
    layer="raw_spliced",
    percent_top=None,
    log1p=False,
    inplace=True,
)
qc_metrics_representation = anndata_string(adata)
print_string_diff(
    text1=latent_time_representation,
    text2=qc_metrics_representation,
    diff_title="Quality control metrics diff",
)
```

```{python}
# | label: post-filtration-plot-genes-by-counts
import seaborn as sns
import pandas as pd

numeric_obs = adata.obs.copy()
numeric_obs["n_genes_by_counts"] = pd.to_numeric(
    numeric_obs["n_genes_by_counts"],
    errors="coerce",
)
qc_counts = sns.jointplot(
    data=numeric_obs,
    x="total_counts",
    y="n_genes_by_counts",
    color="#ff6a14",
    marginal_ticks=True,
    kind="scatter",
    alpha=0.4,
)
qc_counts.plot_joint(
    sns.kdeplot,
    color="gray",
    alpha=0.6,
)
qc_counts.set_axis_labels(
    xlabel="Total counts in cell",
    ylabel="Number of genes >=1 count in cell",
)
```

### Splice state

We can plot the raw data to ensure we have a general understanding of its
sparsity, which may be associated with quality.
We can also plot the raw spliced and unspliced counts.

```{python}
# | label: plot-raw-data
from pyrovelocity.plots import plot_spliced_unspliced_histogram

chart = plot_spliced_unspliced_histogram(
    adata=adata,
    spliced_layer="raw_spliced",
    unspliced_layer="raw_unspliced",
    min_count=3,
    max_count=200,
    title="",
    plot_width=700,
    plot_height=700,
)
chart.save("us_histogram.pdf")
chart
```
